From Reversible Programs to Univalent Universes and Back

### 2017-05-03 upi.tex
started writing

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a univalent (in the sense of Homotopy Type
  Theory) universe. The correspondence relates combinators witnessing type
  isomorphisms in the programming language to paths in the univalent universe,
  and combinator optimizations in the programming language to 2-paths in the
  univalent universe. The result provides a simple computational interpretation
  of paths, 2-paths, and univalence in terms of simple programming
  constructs.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Story and Conjecture}

@@

>>>>

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a univalent (in the sense of Homotopy Type
  Theory) universe. The correspondence relates combinators witnessing type
  isomorphisms in the programming language to paths in the univalent universe,
  and combinator optimizations in the programming language to 2-paths in the
  univalent universe. The result provides a simple computational interpretation
  of paths, 2-paths, and univalence in terms of familiar programming constructs.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A Simple Reversible Programming Language}

The practice of programming languages is replete with \emph{ad hoc} instances of
reversible computations such as database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility but intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961, PhysRevA.32.3266,Toffoli:1980,
  bennett1985fundamental,Frank:1999:REC:930275} and by the potential for
efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.

The early developments of reversible programming languages started with a
conventional programming language, e.g., an extended λ-calculus, and either
extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
imposed constraints on the conventional control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming requires a fresh approach and should be
designed from first principles without the detour via conventional irreversible
languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC, abramsky2005structural,
  DiPierro:2006:RCL:1166042.1166047}.

\begin{verbatim}
Pi... isos, recursion, finite types, Turing complete... nice blah popl 12

ESOP shows a natural 2-layer structure

Examples

conjecture it is related to HoTT

study that language and formalize the connections between its combinators and
between HoTT

stud the smallest core that can explain the idea: restrict to just B; essence
not losing anything really
\end{verbatim}
